Mutation loads in rheumatoid arthritis data from SComatic

### 2024-01-30 scomatic/reports/driver_coverage.Rmd
add driver preprocessing script to find exons, will eventually add to driver_coverage.Rmd, fix infercnv output pattern matching

Genes of interest ('drivers') in rheumatoid arthritis are loaded. Their coverage in the dataset, across samples and cell types, will be checked.

```{r drivers, warning=F, message=F}
drivers <-
  readr::read_tsv('data/driver_genes/driver_gene_coords_for_coverage.tsv')
knitr::kable(drivers)
```

@@

We use `samtools depth` to get the coverage of each site in the genes of interest, both across each sample and within each cell type in each sample.

```{bash cov, eval=F}
# activate conda env
mamba activate jupy

>>>>

Genes of interest in rheumatoid arthritis (hereafter referred to as 'drivers') are defined as genes that are found to undergo significant selection (q-value < 0.2) according to the whole exome and targeted NanoSeq data. The coverage of these sites in the dataset, across samples and celltypes, will be checked.

First, we read in the dNdS results and filter by q-value.

```{r read_dnds, warning = F, message = F}
# read in dnds results
muts <-
  'data/dnds_nanoseq/' %>%
  list.files(pattern = 'nanoseq') %>%
  purrr::map(function(assay) {
    list.files(paste0('data/dnds_nanoseq/', assay), 
               pattern = '_dnds_sel_cv.tsv', 
               full.names = T) %>%
      purrr::map(function(file) {
        file %>%
          readr::read_tsv(show_col_types = F) %>%
          dplyr::mutate(
            assay = assay,
            celltype = file %>% 
              basename() %>% 
              gsub('_dnds_sel_cv.tsv', '', .))
      }) %>%
      dplyr::bind_rows()
  }) %>%
  dplyr::bind_rows() %>%
  dplyr::select(assay, celltype, everything())

# get genes with strong evidence (q-value)
drivers <-
  muts %>%
  dplyr::filter(qglobal_cv < 0.2) %>%
  dplyr::rename(gene = gene_name)

# manually write aliases for genes that are not matched
# HIST1H3B -> H3C2
# HIST1H1E -> H1-4
# CRIPAK   -> This record has been withdrawn by NCBI, after discussions with 
#             CCDS collaborators. It was decided that this locus is not an 
#             independent gene.
drivers <-
  drivers %>%
  dplyr::mutate(
    gene_fixed = dplyr::case_when(
      gene == 'HIST1H3B' ~ 'H3C2',
      gene == 'HIST1H1E' ~ 'H1-4',
      TRUE ~ gene
    )
  )
```

```{r plot_drivers, class.source = "fold-hide", warning = F, message = F}
drivers %>%
  dplyr::distinct(celltype, gene, assay, qglobal_cv) %>%
  dplyr::mutate(neglog10q = -log10(qglobal_cv)) %>%
  ggplot(aes(x = tidytext::reorder_within(gene, -neglog10q, celltype), 
             y = neglog10q, fill = celltype)) +
  geom_col(position = 'dodge') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0)) +
  facet_grid(assay ~ celltype, scales = 'free_x', space = 'free_x') +
  tidytext::scale_x_reordered() +
  labs(x = 'gene', y = '-log10(p)')
```

```{r drivers_df, echo = F}
drivers %>%
  knitr::kable("html") %>% 
  kableExtra::kable_styling("striped", full_width = F) %>% 
  kableExtra::scroll_box(width = "100%", height = "300px")
```

In order to calculate their coverage in the BAMs, we must get their positions.

```{r get_driver_pos}
# load biomart
library(biomaRt)

# load ref
ensembl <- useEnsembl(biomart = 'ensembl',
                      dataset = 'hsapiens_gene_ensembl')

# get positions
drivers_pos <-
  getBM(attributes = c('hgnc_symbol', 'chromosome_name', 'start_position', 'end_position'),
        filters = 'hgnc_symbol', values = unique(drivers$gene),
        mart = ensembl) %>%
  # fix colnames
  dplyr::mutate(gene = hgnc_symbol, 
                chr = chromosome_name, 
                start = start_position,
                end = end_position) %>%
  # remove weird chromosomes
  dplyr::filter(chr %in% c(as.character(1:22, 'X', 'Y')))

# save positions
drivers_pos %>% 
  dplyr::group_by(gene) %>%
  dplyr::summarise(coords = 
                     paste0(unique(chromosome_name), ':', 
                            min(start_position), '-', 
                            max(end_position))) %>%
  readr::write_tsv('data/driver_genes/driver_gene_coords_for_coverage.tsv')
```

@@

We use `samtools depth` to get the coverage of each site in the drivers, both across each sample and within each cell type in each sample.

```{bash cov, eval = T}
# activate conda env
mamba activate jupy
